Making Higher-Order Superposition Work

### 2022-03-24 scratch/paper.tex
intro

\begin{abstract}%
  The automatic discharge of tedious subgoals is high on the wishlist of many
  users of proof assistants. Some proof assistants discharge such goals
  by translating them to first-order logic and invoking an efficient prover on
  them, but much is lost in translation. As an alternative,
  we propose to extend first-order provers with native support for
  higher-order features. Building on our extension of E to $\lambda$-free
  higher-order logic, we now extend E to full higher-order logic.
  The resulting prover is the second strongest on both TPTP benchmarks and
  benchmarks coming from a proof assistant.
  %It also incurs no overhead on first-order problems.  -- sounds like a detail --JB
\end{abstract}

@@

\section{Introduction}
\label{sec:introduction}

@@

In this \paper, we fulfill a three-year-old promise: We present the extension of
Ehoh to full higher-order logic (Sect.~\ref{sec:logic}) using incomplete variants
of $\lambda$-superposition. We call this prover \ehohii.
%
The $\lambda$-superposition calculi were
previously implemented in Zipperposition, and
extensive experiments with various heuristic choices have been performed
\cite{section-making-ho-work}. In \ehohii{}'s implementation, we used
these experiences to choose a set of effective rules that could easily be
retrofitted into an originally first-order prover. Another principle that guided 
the design of \ehohii{} was \emph{gracefulness}: we made sure that our changes
do not impact already strong first-order performance of Ehoh.

An simpler approach to support higher-order logic is to implement
combinatory superposition, which is based on \textsf{SKBCI}-combinators
and implemented in Vampire \cite{br-20-full-sup-w-combs}.
These can be supported as a thin layer on top of $\lambda$-free terms. However,
in unpublished experiments, we found that $\lambda$-superposition performs
substantially better, corroborating the CASC results, so we
decided to make a more profound change to Ehoh and implement
$\lambda$-superposition.

@@

One of the main challenges we faced was retrofitting $\lambda$-terms in Ehoh's
term representation (Sect.~\ref{sec:terms}). Furthermore, Ehoh's main inference
engine was designed with the assumption that it will be used with
inferences that compute a most general unifier. We
implementd a higher-order unification procedure \cite{unif-section}
that can return multiple unifiers (Sect.~\ref{sec:unif-match-index}) and
integrated it in the inference engine. Finally, we extended and adapted the
superposition rule, resulting in an incomplete, pragmatic variant of
$\lambda$-superposition (Sect.~\ref{sec:calculus}).

% most likely text
We \NumberOK{evaluated} \ehohii{} on a selection of proof assistants benchmarks
as well as all higher-order theorems in the TPTP library \cite{gs-17-tptp}
(Sect.~\ref{sec:eval}). We found
that \ehohii{} clearly outperforms Ehoh on all benchmarks. Compared
with other higher-order provers, it is second
only to the cooperative version of Zipperposition, which employs Ehoh as a
backend. A fairer comparison without the backend puts \ehohii{} in the
first place for both benchmark suites.
We also compared the performance of \ehohii{} with E on first-order
%%% @PETAR: I made this slightly sentence shorter and more forceful. Please
%%% check if you agree. --JB
problems and found that little overhead has been introduced by the
extension to higher-order logic.

@@

\section{Logic} %---0.5p
\label{sec:logic}

@@

We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$ and
$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$.
Every $\beta$-normal term can be written as $\lam{\overline{x}_m}{s \,
\overline{t}_n}$, where $s$ is not an application; we call $s$ the \emph{head}
of the term. If $s$ is a free variable, we call the term \emph{flex};
otherwise, the term is \emph{rigid}. If the type of a term is of the form $\tau_1 \to \cdots \to
\tau_n \to o$, where $o$ is the distinguished Boolean type and $n \ge 0$, we
call it a \emph{predicate}. A literal $l$ is an equation $s \eq t$ or
a disequation $s \not\eq t$. A clause is a finite multiset of literals,
interpreted and written disjunctively:\ $l_1 \llor \cdots \llor l_n$.
Logical symbols that may occur within terms are written in bold:
$\pmb{\neg}, \iand, \ior, \iimplies, \iequiv, \dots$. Predicate
literals are encoded as (dis)equations with $\itrue$ based on their sign; for
example, $\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$.

\section{Terms}
\label{sec:terms}

Ehoh's term representation is centered around perfect term sharing %or exa
\cite{ls-01-shared}: Any two structurally identical terms are
guaranteed to be the same object in memory. This is achieved through term
\emph{cells}, which represent individual terms. Each cell has (among other fields)
(1)~\texttt{f\_code}, an integer corresponding to the symbol at the head of the term (negative
if the head is a free variable, positive otherwise); (2)~\texttt{num\_args},
corresponding to the number of arguments applied to the head; and (3)~\texttt{args},
a size-\texttt{num\_args} array of pointers to argument terms. We
use the first-order notation $\cst{f}(s_1, \ldots, s_n)$ to denote a cell whose
\texttt{f\_code} corresponds to $\cst{f}$, \texttt{num\_args} equals $n$, and
\texttt{args} points to the cells for $s_1, \ldots s_n$.

@@

Other than possibly being instantiated during $\beta$-reduction, De Bruijn
indices mostly behave as constants. Therefore, we decided to represent De
Bruijn indices using positive \texttt{f\_code}s: De Bruijn index of value $i$
will have $i$ as the \texttt{f\_code}. To ensure De Bruijn indices are not
mistaken for function symbols, we use the \texttt{properties} bitfield of the
cell, which holds precomputed properties of %shared cell
%%% @PETAR(checked, you are right): I don't understand why you wrote "shared cell". Please double-check. --JB
the cell. We introduce the
property \texttt{IsDBVar} to denote that the cell represents a De
Bruijn index. Any attempt to create a De Bruijn index is performed through
a dedicated library function that sets \texttt{IsDBVar} property for every term it
returns. When given the same De Bruijn index and type, this function is
guaranteed to always return the same object. Finally, we have guarded all the
functions and macros that manipulate function codes with the check if the
property \texttt{IsDBVar} is set. To ensure perfect sharing of every occurrence
of De Bruijn indices, arguments to De Bruijn indices are applied like for free
variables, using \internalat{}.

@@

To avoid this, we use Ehoh's perfectly shared
terms and their \texttt{properties} field.
%
We introduce the property \texttt{HasBetaReducibleSubterm} that indicates whether
%%% @PETAR(is is good): "is" or "may be" below? (Overapproximation?) --JB
a cell is $\beta$-reducible.
Whenever a new cell that contains a
$\beta$-reducible term as a direct subterm is shared, the property is set.
Setting of the property is inductively continued when further superterms are
shared. For example, in the term $t = \cst{f} \, \cst{a} \, (\cst{g} ((\lambda x.\,
x)\,\cst{a}))$, the cells for $(\lambda x.\, x)\,\cst{a}$,
$\cst{g}\,((\lambda x.\, x)\,\cst{a})$, and $t$ itself have the property
\texttt{HasBetaReducibleSubterm} set.
%
When it needs to find $\beta$-reducible subterms, \ehohii{} will visit only the
cells with this property set. This further means that on first-order
subterms, a single bit masking operation is enough to determine that no subterm
should be visited.

@@

\looseness=-1
More specifically, when $\eta$-reducing a cell $\internallam(\db{0}, s)$,
\ehohii{} considers all $\lambda$ binders in $s$ as well. In general,
the cell will be of the form
$\internallam(\db{0}, \dotsc,\allowbreak \internallam(\db{0}, t) \ldots)$,
where $t$ is not a
$\lambda$-abstraction. Then \ehohii{} breaks down the body~$t$ into a maximal
decomposition $u\, (\dbvar{n}\db{{}-1}) \, \ldots \, \db{1} \, \db{0}$.
If $n = 0$, the cell is not $\eta$-reducible.
Otherwise, $u$ is traversed to determine the
minimal index~$\dbvar{j}$ of a loose De Bruijn index,
taking $\dbvar{j} = \infty$ if no such index exists.
%%% @PETAR(I prefer leftmost as we are going top level left-to-right and not
%bottom-up as innermost can suggest): Check "innermost".
%%% @PETAR(you are right, but I will add righmost outermost to be more precise): But it's rightmost then not leftmost! In the example, the very first
%%% binder remains (the one to which the 2's point). --JB
We can then remove the $k = \min(j,n)$ rightmost outermost $\lambda$ binders in $\internallam(\db{0},
\ldots, \internallam(\db{0}, t) \ldots)$ and replace %the body
$t$ by the variant of
$u \allowbreak\, (\dbvar{n}\db{{}-1}) \allowbreak\, \ldots \, (\dbvar{k}\db{{}+1}) \, \dbvar{k}$
obtained by shifting the loose De Bruijn indices down by~$k$.

@@

\section{Unification, Matching, and Term Indexing}
\label{sec:unif-match-index}
% \begin{itemize}
%   \item Procedure that implements a variant of the algo described in Efficient
%   HO Unif paper.
%   \item Two oracles -- fixpoint and pattern unif (solid not implemented as
%   complexity/gain ratio is quite negative)
%   \item It is implemented iteratively, with subtle optimizations
%   \item Its iterator is completely contained -- iterators can be stored and
%   later waked up similarly to what we do in Zipperposition.
%   \item Mention that so far the prover fares extremely well with complicated
%   unification problems and that we do not consider implementing more complex
%   given clause algo in Ehoh.
%   \item Matching works only with algorithms that return a single matcher:
%   pattern matching and LFHOL matching : main optimization from E is preserved.
%   PDT algorithm also works with both kinds of terms LFHOL and pattern. Check if
%   something is a pattern is efficient!
%   \item Fingerprints and feature vectors as in Zipperposition.
%   \item To avoid huge refactorings iterator is used even if the problem to be
%   solved is first-order.  
% \end{itemize}

Standard superposition crucially depends on the concept of a most general unifier
(MGU).
In higher-order logic, such a unifier does not in general exist, and the concept
%%% @PETAR: CSU, infinite
is replaced by that of a complete set of unifiers (CSU), which may be infinite.
Vukmirovi\'c et al.\ \cite{unif-section} designed an efficient
procedure to enumerate a CSU for a term pair. They
implemented it in Zipperposition, together with some extensions to term indexing.
In \ehohii{}, we further improve the performance of this procedure by implementing a terminating,
incomplete variant. We also introduce a new indexing data structure.

@@

\label{sec:calculus}
% \begin{itemize}  
  % \item The main goal is simplicity, efficiency and maximal compatibility with
  %     FO superposition.
  % \item More or less the same as LFHOL sup, but now computes CSUs instead
  %     of LFHOL MGUs (and (potentially) uses new orders).
  % \item Out of Boolean rules we have Boolean simplification and black-box clausification.
  %     Most importantly -- we have no Boolean hoisting rules : everything is dealt with
  %     ext-sup, and FOOL clausification! It simply works as good as Boolean hoisting.
  % \item IMPORTANT: Prefix optimization is disabled, now we rely on ArgCong and NegExt.
  % \item Many extensions:  ExtSup (With the same indexing), leibniz elimination, predicate enumeration,
  %     FlexElim, Choice instantiation, recognition of induction.
  % \end{itemize}

@@

\ourpara{Calculus} \ehohii{} implements the same superposition
calculus as Ehoh with three important changes. First, wherever Ehoh requires
the most general unifier of terms, \ehohii{} enumerates unifiers from a finite subset of the CSU, as explained in
Sect.~\ref{sec:unif-match-index}. Second, \ehohii{} uses versions of the KBO and
RPO orders designed for $\lambda$-terms.
% as described by Bentkamp et al.\ \cite{bbtv-21-full-ho-sup}.

%TODO: @JASMIN: How do we cite the STTT article?
Third difference is more subtle. One of the main features of Ehoh is
\emph{prefix optimization} \cite[Sect.~1]{section-ehoh}: a method which, given a demodulator $s \eq t$, allows
to replace both applied and unapplied occurrences of $s$ by $t$ 
by traversing only first-order subterms of a rewritable term. In $\lambda$-free setting this
optimization is useful, but in presence of $\beta\eta$-normalization, shapes of
terms can change drastically, making it much harder to
track prefixes of terms. This is why we disabled the prefix optimization in \ehohii{}.
To counterbalance this removal, we introduced
the argument congruence rule \infname{AC} %(also called \infname{ArgCong})
in \ehohii{}
and enabled positive and negative functional extensionality (\infname{PE} and
\infname{NE}) by default:

@@

\section{Evaluation}
\label{sec:eval}

In this section, we try to answer two questions about \ehohii{}: \emph{How does
\ehohii{} compare against other higher-order provers (including Ehoh)?} \emph{Does
\ehohii{} introduce any overhead compared with Ehoh?} To answer
these questions, we ran provers on problems from the TPTP library \cite{gs-17-tptp} and on
benchmars generated by Sledgehammer (SH) \cite{pb-12-sh}.
The experiments were carried out on StarExec Miami
\cite{sst-14-starexec} nodes equipped with Intel Xeon E5-2620 v4 CPU clocked at
2.10~GHz. For the TPTP part, we used the CASC-28\footnote{\url{http://www.tptp.org/CASC/28/}} time limits:\ 120~s
wallclock and 960~s CPU. For SH, we used Sledgehammer's
default time limit:\ 30~s wallclock and CPU.

@@

\ehohii{} dramatically improves the higher-order reasoning capabilities
compared to Ehoh. It solves \NumberOK{20\%} more problems on TPTP benchmarks
and \NumberOK{3\%} more problems on SH benchmarks, where Ehoh was already very
successful.

\ehohii{} also outperforms all other higher-order provers other than Zipperposition-coop.
If Zipperposition's Ehoh backend is disabled,
\ehohii{} outperforms Zipperposition by a wide margin.
This comparison is arguably fairer; after all,
\ehohii{} does not use an older version of Zipperposition
as a backend.
%
These results suggest that \ehohii{} already implements most of the necessary features
for a high-performance higher-order prover
but could benefit from the kind of fine-tuning that
Zipperposition underwent in the last two years.

Remarkably, there is a substantial disparity in the set of problems solved by
\ehohii{} and Zipperposition-coop. The raw data show thats \ehohii{} solves
\NumberOK{144} SH problems and \NumberOK{24} TPTP problems that Zipperposition-coop
does not. The lower number of uniquely solved TPTP problems is most likely 
due to Zipperposition's being heavily optimized on TPTP.

\ourpara{Comparison with the First-Order E} Both Ehoh and \ehohii{} can be compiled
in a mode that disables most of the higher-order reasoning. This mode is
designed for users that are interested only in E's first-order capabilities and
care a lot about performance.
To answer the second
evaluation question, about assessing overhead of \ehohii{},
we chose all the \NumberOK{1138} unique problems used at CASC
from 2019 to 2021 in first-order theorem division and ran Ehoh and \ehohii{}
both in this first-order (FO) mode and in higher-order (HO) mode.

@@

                            & TPTP       & SH        \\
      \midrule
      cvc5                  & 1931       & 2577      \\
      Ehoh                  & 2105       & 2611      \\
      \ehohii{}             & 2527       & 2699      \\
      Leo-III               & 2282       & 1601      \\
      Satallax              & 2320       & 1719      \\
      Vampire               & 2203       & 2240      \\
      Zipperposition-coop   & {\bf 2583} & {\bf 2754}\\
      Zipperposition-uncoop & 2483       & 2181      \\
      \bottomrule
    \end{tabular}}
    \caption{Comparison of higher-order provers}
    \label{fig:all-provers}
  \end{minipage}\hfill\hfill
  \begin{minipage}[ht!]{.30\linewidth}
    \centering
    \def\arraystretch{1.1}%
    \relax{\begin{tabular}{l@{\hskip 1em}@{\hskip 0.5em}c} \toprule
                      & TPTP     \\
      \midrule
      Ehoh--FO      & 535      \\
      Ehoh--HO      & 538      \\
      \ehohii{}--FO & 533      \\
      \ehohii{}--HO & {\bf 541}\\
      \bottomrule
    \end{tabular}}
    \caption{Evaluation of \ehohii{}'s overhead}
    \label{fig:ehoh}
  \end{minipage}\hfill\hfill
\end{figure}

\section{Discussion and Related Work} %---0.5p
\label{sec:discussion}

@@

Extending first-order provers with higher-order reasoning capabilities has been
tried by other researchers as well. Bhayat and Reger first extended Vampire to
higher-order logic using combinatory unification \cite{br-20-full-sup-w-combs},
an incomplete approach, before they designed an implemented a complete
higher-order superposition calculus based on combinators \cite{br-19-restricted-unif}.
This calculus is also implemented in Zipperposition. Barbosa et al.\ extended
the SMT solvers CVC4 (now cvc5) and veriT to higher-order logic in an incomplete
way \cite{brotb-19-ho-smt}.

@@

\section{Conclusion} %---0.5p
\label{sec:conclusion}

>>>>

\begin{abstract}%
  The automatic discharge of tedious subgoals is high on the wishlist of many
  users of proof assistants. Some proof assistants discharge such goals
  by translating them to first-order logic and invoking an efficient prover on
  them, but much is lost in translation. As an alternative,
  we propose to extend first-order provers with native support for
  higher-order features. Building on our extension of E to $\lambda$-free
  higher-order logic, we now extend E to full higher-order logic.
  The resulting prover is the \NumberOK{strongest} one on benchmarks coming from a
  proof assistant, and the second best on TPTP benchmarks.
  %It also incurs no overhead on first-order problems.  -- sounds like a detail --JB
\end{abstract}

@@

\section{Introduction}
\label{sec:ehoh2:introduction}

@@

In this chapter, we fulfill a three-year-old promise: We present the extension of
Ehoh to full higher-order logic (Sect.~\ref{sec:ehoh2:logic}) using incomplete variants
of $\lambda$-superposition. We call this prover \ehohii.
%
The $\lambda$-superposition calculi were
previously implemented in Zipperposition, and
extensive experiments with various heuristic choices have been performed
\cite{section-making-ho-work}. In \ehohii{}'s implementation, we used
these experiences to choose a set of effective rules that could easily be
retrofitted into an originally first-order prover. Another principle that guided 
the design of \ehohii{} was \emph{gracefulness}: we made sure that our changes
do not impact the strong first-order performance of E and Ehoh.

@@

One of the main challenges we faced was retrofitting $\lambda$-terms in Ehoh's
term representation (Sect.~\ref{sec:ehoh2:terms}). Furthermore, Ehoh's main inference
engine was designed with the assumption that it will be used with
inferences that compute a most general unifier. We
implementd a higher-order unification procedure \cite{unif-section}
that can return multiple unifiers (Sect.~\ref{sec:ehoh2:unif-match-index}) and
integrated it in the inference engine. Finally, we extended and adapted the
superposition rule, resulting in an incomplete, pragmatic variant of
$\lambda$-superposition (Sect.~\ref{sec:ehoh2:calculus}).

We \NumberOK{evaluated} \ehohii{} on a selection of proof assistants benchmarks
as well as all higher-order theorems in the TPTP library \cite{gs-17-tptp}
(Sect.~\ref{sec:ehoh2:eval}). We found
that \ehohii{} clearly outperforms Ehoh on all benchmarks. It outperformed all other higher-order provers on
proof assistant benchmarks; on TPTP benchmarks it ended up second only to 
to the cooperative version of Zipperposition, which employs Ehoh as a
backend. An arguably fairer comparison without the backend puts \ehohii{} in the
first place for both benchmark suites.
We also compared the performance of \ehohii{} with E on first-order
%%% @PETAR: I made this slightly sentence shorter and more forceful. Please
%%% check if you agree. --JB
problems and found that little overhead has been introduced by the
extension to higher-order logic.

@@

\section{Logic} %---0.5p
\label{sec:ehoh2:logic}

@@

We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$ and
$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. Every
$\beta$-normal term can be written as $\lam{\overline{x}_m}{s \,
\overline{t}_n}$, where $s$ is not an application; we call $s$ the \emph{head}
of the term. If $s$ is a free variable, we call the term \emph{flex}; otherwise,
the term is \emph{rigid}. A term of type $o$, where $o$ is the distinguished Boolean type,
is called a \emph{formula}. A type whose type is of the form $\tau_1 \to
\cdots \to \tau_n \to o$ is called
a \emph{predicate}. Logical symbols are part of the signature and
may thus occur within terms. We write them in bold: $\pmb{\neg}, \iand, \ior,
\iimplies, \iequiv, \ieq, \dots$.

On top of the terms, we define some clausal structure, as is customary in
first-order logic.
A literal $l$ is an equation $s \eq t$ or a
disequation $s \not\eq t$. A clause is a finite multiset of literals,
interpreted and written disjunctively:\ $l_1 \llor \cdots \llor l_n$.
Notice that the clause-level operators are not set in bold.
Predicate literals are encoded as (dis)equations with $\itrue$ based on their sign; for
example, $\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ as $\neglit{\cst{even}(x)}$.

\section{Terms}
\label{sec:ehoh2:terms}

\looseness=-1
E has been designed around perfect term sharing %or exa
\cite{ls-01-shared}, a design that we carried on to  Ehoh and \ehohii{}: Any two structurally identical terms are
guaranteed to be the same object in memory. This is achieved through term
\emph{cells}, which represent individual terms. Each cell has (among other fields)
(1)~\texttt{f\_code}, an integer corresponding to the symbol at the head of the term (negative
if the head is a free variable, positive otherwise); (2)~\texttt{num\_args},
corresponding to the number of arguments applied to the head; and (3)~\texttt{args},
a size-\texttt{num\_args} array of pointers to argument terms. We
use the first-order notation $\cst{f}(s_1, \ldots, s_n)$ to denote a cell whose
\texttt{f\_code} corresponds to $\cst{f}$, \texttt{num\_args} equals $n$, and
\texttt{args} points to the cells for $s_1, \ldots s_n$.

@@

Other than possibly being instantiated during $\beta$-reduction, De Bruijn
indices mostly behave as constants. Therefore, we decided to represent De
Bruijn indices using positive \texttt{f\_code}s: The De Bruijn index of value $i$
will have $i$ as the~\verb|f_code|. To ensure De Bruijn indices are not
mistaken for function symbols, we use the \texttt{properties} bitfield of the
cell, which holds precomputed properties of %shared cell
%%% @PETAR(checked, you are right): I don't understand why you wrote "shared cell". Please double-check. --JB
the cell. We introduce the
property \texttt{IsDBVar} to denote that the cell represents a De
Bruijn index. Any attempt to create a De Bruijn index is performed through
a dedicated library function that sets \texttt{IsDBVar} property for every term it
returns. When given the same De Bruijn index and type, this function is
guaranteed to always return the same object. Finally, we have guarded all the
functions and macros that manipulate function codes with the check if the
property \texttt{IsDBVar} is set. To ensure perfect sharing of every occurrence
of De Bruijn indices, arguments to De Bruijn indices are applied like for free
variables, using \internalat{}.

@@

To avoid this, we use Ehoh's perfectly shared
terms and their \texttt{properties} field.
%
We introduce the property \texttt{HasBetaReducibleSubterm} which is set if
%%% @PETAR(is is good): "is" or "may be" below? (Overapproximation?) --JB
a cell is $\beta$-reducible.
Whenever a new cell that contains a
$\beta$-reducible term as a direct subterm is shared, the property is set.
Setting of the property is inductively continued when further superterms are
shared. For example, in the term $t = \cst{f} \, \cst{a} \, (\cst{g} ((\lambda x.\,
x)\,\cst{a}))$, the cells for $(\lambda x.\, x)\,\cst{a}$,
$\cst{g}\,((\lambda x.\, x)\,\cst{a})$, and $t$ itself have the property
\texttt{HasBetaReducibleSubterm} set.
%
When it needs to find $\beta$-reducible subterms, \ehohii{} will visit only the
cells with this property set. This further means that on first-order
subterms, a single bit masking operation is enough to determine that no subterm
should be visited.

@@

\looseness=-1
More specifically, when $\eta$-reducing a cell $\internallam(\db{0}, s)$,
\ehohii{} considers all $\lambda$ binders in $s$ as well. In general,
the cell will be of the form
$\internallam(\db{0}, \dotsc,\allowbreak \internallam(\db{0}, t) \ldots)$,
where $t$ is not a
$\lambda$-abstraction, and $l$ is the number of $\internallam$ symbols above $t$. Then \ehohii{} breaks down the body~$t$ into a maximal
decomposition $u\, (\dbvar{n}\db{{}-1}) \, \ldots \, \db{1} \, \db{0}$.
If $n = 0$, the cell is not $\eta$-reducible.
Otherwise, $u$ is traversed to determine the
minimal index~$\dbvar{j}$ of a loose De Bruijn index,
taking $\dbvar{j} = \infty$ if no such index exists.
%%% @PETAR(I prefer leftmost as we are going top level left-to-right and not
%bottom-up as innermost can suggest): Check "innermost".
%%% @PETAR(you are right, but I will add righmost outermost to be more precise): But it's rightmost then not leftmost! In the example, the very first
%%% binder remains (the one to which the 2's point). --JB
We can then remove the $k = \min\{j,l,n\}$ rightmost outermost $\lambda$ binders in $\internallam(\db{0},
\ldots, \internallam(\db{0}, t) \ldots)$ and replace %the body
$t$ by the variant of
$u \allowbreak\, (\dbvar{n}\db{{}-1}) \allowbreak\, \ldots \, (\dbvar{k}\db{{}+1}) \, \dbvar{k}$
obtained by shifting the loose De Bruijn indices down by~$k$.

@@

\section{Unification, Matching, and Term Indexing}
\label{sec:ehoh2:unif-match-index}
% \begin{itemize}
%   \item Procedure that implements a variant of the algo described in Efficient
%   HO Unif paper.
%   \item Two oracles -- fixpoint and pattern unif (solid not implemented as
%   complexity/gain ratio is quite negative)
%   \item It is implemented iteratively, with subtle optimizations
%   \item Its iterator is completely contained -- iterators can be stored and
%   later waked up similarly to what we do in Zipperposition.
%   \item Mention that so far the prover fares extremely well with complicated
%   unification problems and that we do not consider implementing more complex
%   given clause algo in Ehoh.
%   \item Matching works only with algorithms that return a single matcher:
%   pattern matching and LFHOL matching : main optimization from E is preserved.
%   PDT algorithm also works with both kinds of terms LFHOL and pattern. Check if
%   something is a pattern is efficient!
%   \item Fingerprints and feature vectors as in Zipperposition.
%   \item To avoid huge refactorings iterator is used even if the problem to be
%   solved is first-order.  
% \end{itemize}

Standard superposition crucially depends on the concept of a most general unifier
(MGU).
In higher-order logic, such a unifier does not in general exist, and the concept
%%% @PETAR: CSU, infinite
is replaced by that of a complete set of unifiers (CSU), which may be infinite.
Vukmirovi\'c et al.\ \cite{unif-section} designed an efficient
procedure to enumerate a CSU for a term pair. It is implemented in Zipperposition, together with some extensions to term indexing.
In \ehohii{}, we further improve the performance of this procedure by implementing a terminating,
incomplete variant. We also introduce a new indexing data structure.

@@

\label{sec:ehoh2:calculus}
% \begin{itemize}  
  % \item The main goal is simplicity, efficiency and maximal compatibility with
  %     FO superposition.
  % \item More or less the same as LFHOL sup, but now computes CSUs instead
  %     of LFHOL MGUs (and (potentially) uses new orders).
  % \item Out of Boolean rules we have Boolean simplification and black-box clausification.
  %     Most importantly -- we have no Boolean hoisting rules : everything is dealt with
  %     ext-sup, and FOOL clausification! It simply works as good as Boolean hoisting.
  % \item IMPORTANT: Prefix optimization is disabled, now we rely on ArgCong and NegExt.
  % \item Many extensions:  ExtSup (With the same indexing), leibniz elimination, predicate enumeration,
  %     FlexElim, Choice instantiation, recognition of induction.
  % \end{itemize}

@@

\ourpara{Calculus} \ehohii{} implements the same superposition
calculus as Ehoh with three important changes. First, wherever Ehoh requires
the most general unifier of terms, \ehohii{} enumerates unifiers from a finite subset of the CSU, as explained in
Sect.~\ref{sec:ehoh2:unif-match-index}. Second, \ehohii{} uses versions of the KBO and
LPO orders designed for $\lambda$-terms.
% as described by Bentkamp et al.\ \cite{bbtv-21-full-ho-sup}.

%TODO: @JASMIN: How do we cite the STTT article?
The third difference is more subtle. One of the main features of Ehoh is
\emph{prefix optimization} \cite[Sect.~1]{section-ehoh}: a method which, given a demodulator $s \eq t$, allows
to replace both applied and unapplied occurrences of $s$ by $t$ 
by traversing only first-order subterms of a rewritable term. In $\lambda$-free setting this
optimization is useful, but in presence of $\beta\eta$-normalization, shapes of
terms can change drastically, making it much harder to
track prefixes of terms. This is why we disabled the prefix optimization in \ehohii{}.
To counterbalance this removal, we introduced
the argument congruence rule \infname{AC} %(also called \infname{ArgCong})
in \ehohii{}
and enabled positive and negative functional extensionality (\infname{PE} and
\infname{NE}) by default:

@@

\section{Evaluation}
\label{sec:ehoh2:eval}

In this section, we try to answer two questions about \ehohii{}: \emph{How does
\ehohii{} compare against other higher-order provers (including Ehoh)?} \emph{Does
\ehohii{} introduce any overhead compared with Ehoh?} To answer
these questions, we ran provers on problems from the TPTP library \cite{gs-17-tptp} and on
benchmarks generated by Sledgehammer (SH) \cite{pb-12-sh}.
The experiments were carried out on StarExec Miami
\cite{sst-14-starexec} nodes equipped with Intel Xeon E5-2620 v4 CPU clocked at
2.10~GHz. For the TPTP part, we used the CASC-28\footnote{\url{http://www.tptp.org/CASC/28/}} time limits:\ 120~s
wallclock and 960~s CPU. For SH and to answer the other question, we used Sledgehammer's
default time limit:\ 30~s wallclock and CPU.

@@

\ehohii{} dramatically improves the higher-order reasoning capabilities
compared to Ehoh. It solves \NumberOK{20\%} more problems on TPTP benchmarks
and \NumberOK{6\%} more problems on SH benchmarks, where Ehoh was already very
successful.

\ehohii{} was mainly designed as efficient backed to proof assistants. As such it 
excels on SH benchmarks. On TPTP, it outperforms all higher-order provers other than Zipperposition-coop.
If Zipperposition's Ehoh backend is disabled,
\ehohii{} outperforms Zipperposition by a wide margin.
This comparison is arguably fairer; after all,
\ehohii{} does not use an older version of Zipperposition
as a backend.
%
These results suggest that \ehohii{} already implements most of the necessary features
for a high-performance higher-order prover
but could benefit from the kind of fine-tuning that
Zipperposition underwent in the last two years.

Remarkably, there is a substantial disparity in the set of problems solved by
\ehohii{} and Zipperposition-coop. The raw data show thats \ehohii{} solves
\NumberNOK{166} SH problems and \NumberNOK{21} TPTP problems that Zipperposition-coop
does not. The lower number of uniquely solved TPTP problems is most likely 
due to Zipperposition's being heavily optimized on TPTP.

\ourpara{Comparison with the First-Order E} Both Ehoh and \ehohii{} can be compiled
in a mode that disables most of the higher-order reasoning. This mode is
designed for users that are interested only in E's first-order capabilities and
care a lot about performance.
To answer the second
evaluation question, about assessing overhead of \ehohii{},
we chose all the \NumberNOK{1138} unique problems used at CASC
from 2019 to 2021 in first-order theorem division and ran Ehoh and \ehohii{}
both in this first-order (FO) mode and in higher-order (HO) mode.

@@

                            & TPTP       & SH        \\
      \midrule
      cvc5                  & 1931       & 2577      \\
      Ehoh                  & 2105       & 2611      \\
      \ehohii{}             & 2523       & {\bf 2766}\\
      Leo-III               & 2282       & 1601      \\
      Satallax              & 2320       & 1719      \\
      Vampire               & 2203       & 2240      \\
      Zipperposition-coop   & {\bf 2583} & 2754      \\
      Zipperposition-uncoop & 2483       & 2181      \\
      \bottomrule
    \end{tabular}}
    \caption{Comparison of higher-order provers}
    \label{fig:all-provers}
  \end{minipage}\hfill\hfill
  \begin{minipage}[ht!]{.30\linewidth}
    \centering
    \def\arraystretch{1.1}%
    \relax{\begin{tabular}{l@{\hskip 1em}@{\hskip 0.5em}c} \toprule
                      & TPTP     \\
      \midrule
      Ehoh -- FO      & 535      \\
      Ehoh -- HO      & 538      \\
      \ehohii{} -- FO & {\bf 539}\\
      \ehohii{} -- HO & {\bf 539}\\
      \bottomrule
    \end{tabular}}
    \caption{Evaluation of \ehohii{}'s overhead}
    \label{fig:ehoh}
  \end{minipage}\hfill\hfill
\end{figure}

\section{Discussion and Related Work} %---0.5p
\label{sec:ehoh2:discussion}

@@

Extending first-order provers with higher-order reasoning capabilities has been
attempted by other researchers as well.
Barbosa et al.\ extended
the SMT solvers CVC4 (now cvc5) and veriT to higher-order logic in an incomplete
way \cite{brotb-19-ho-smt}.
Bhayat and Reger first extended Vampire to
higher-order logic using combinatory unification \cite{br-20-full-sup-w-combs},
an incomplete approach, before they designed and implemented a complete
higher-order superposition calculus based on \textsf{SKBCI}-com\-binators \cite{br-19-comb-unif}.
The advantage is that combinators can be supported as a thin layer on top of
$\lambda$-free terms.
This calculus is also implemented in Zipperposition.
However, in unpublished experiments, we found that $\lambda$-superposition
performs substantially better, corroborating the CASC results, so we decided to
make a more profound change to Ehoh and implement $\lambda$-superposition.

@@

\section{Conclusion} %---0.5p
\label{sec:ehoh2:conclusion}

In 2019, the reviewers of our paper introducing Ehoh \cite{section-ehoh} were
skeptical that extending it with support for full higher-order logic would be
feasible. One of them wrote:
%
\begin{quote}
A potential criticism could be that this step from E to Ehoh is just extending
FOL by those aspects of HOL that are easily in reach with rather straightforward
extensions (none of the extensions is indeed very complicated), and that the
difficult challenges of fully supporting HOL have yet to be confronted.
\end{quote}
%
We ended up addressing the theoretical ``difficult challenges'' in other work
with colleagues. In this chapter, we faced the practical challenges pertaining to the extension
of Ehoh's data structures and algorithms to support full higher-order logic and
demonstrated that such an extension is possible. \NumberOK{Our evaluation} shows
that this extension makes \ehohii{} the best higher-order prover on
benchmarks coming from interactive theorem proving practice, which has been our goal since 2017. \ehohii{}'s
performance lags slightly behind Zipperposition on TPTP problems, possibly because Zipperposition
does not assume a clausal structure and can perform subtle formula-level
inferences. In the future, we plan to implement the same features in \ehohii{}.
We have also only started tuning \ehohii{}'s heuristics on higher-order
problems.
%%% @PETAR(read): By submission time, CASC will already have taken place!
% We plan to train the prover better. ahead of the next CASC competition.
